---
title: "Charcoal Aerial Dispersion and Source Area Model"
subtitle: "Watershed V.1"
author: "Grant Snitker - Arizona State University"
date: "12/8/2016"
output: html_document
---
### Model Description
This R script adapts Clark's (1988) 2-d model of primary charcoal dispersion into a 3-d source area model for all cells located within a watershed.Charcoal assemblages recovered from the lower reaches of a watershed represent cumulative fire activity and vegetation located upstream of the sample location. These materials will have moved downstream due to channel flow, sheet wash, and other fluvial processes until they were periodically deposited during a high flow or flood event.

### Set-up and Parameters

##### *Set working directory and libraries*
```{r Libraries, , message=FALSE}
#setwd("Set working directory") # change to personal working directory
library(raster)
library(rgeos)
library(pracma)
library(plyr)
library(rgdal)
library(parallel)
library(ggmap)
library(ggplot2)
library(maptools)
```

##### *Load required raster inputs, resample rasters to desired resolution, and select cells analysis*
```{r Download Data, ,message=FALSE}
### Download all required rasters from github
download.file(url = "https://raw.githubusercontent.com/gsnitker/Charcoal_Source_Area/master/Valencia_DEM.tif", destfile = "./Valencia_DEM.tif", method = "auto") # elevation raster

download.file(url = "https://raw.githubusercontent.com/gsnitker/Charcoal_Source_Area/master/navarres_50mres_wind_dir.tif", destfile = "./navarres_50mres_wind_dir.tif", method = "auto") # wind direction raster

download.file(url = "https://raw.githubusercontent.com/gsnitker/Charcoal_Source_Area/master/navarres_50mres_wind_speed.tif", destfile = "./navarres_50mres_wind_speed.tif", method = "auto") # wind velocity raster

download.file(url = "https://raw.githubusercontent.com/gsnitker/Charcoal_Source_Area/master/SP_NV_5.zip", destfile = "./SP_NV_5.zip", method = "auto") # watershed shapefile

### Load downloaded data
elev <- raster("Valencia_DEM.tif") # elevation
wind_dir = raster("navarres_50mres_wind_dir.tif") # wind direction
wind_vel = raster("navarres_50mres_wind_speed.tif") # wind velocity
unzip("SP_NV_5.zip")
sample.poly = shapefile("./SP_NV_5/SP_NV_5_Watershed.shp")

### Re-project and re-sample data to desired resolution (in this case, resmaple to raster with the lowest resolution)
#Re-project
crs(elev) = "+proj=utm +zone=30 +datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0"
#Re-sample
elev = resample(elev, wind_vel, method='bilinear')# re-sample rasters to the same resolution
wind_dir = resample(wind_dir, wind_vel, method='bilinear') # re-sample rasters to the same resolution
# reverse wind direction and keep within 360 degree scale
wind_dir_trans_1 = wind_dir - 180 # reverse values
wind_dir_trans_2 = wind_dir - 180 # reverse values
wind_dir_trans_1[wind_dir_trans_1 <= 0] = NA #select positive values that need no more transformation 
wind_dir_trans_2[wind_dir_trans_2 >= 0] = NA #select negative values that need more transformation 
wind_dir_trans_2 = wind_dir_trans_2 + 360 # add 360 to values that were negative 
wind_dir_trans = merge(wind_dir_trans_1,wind_dir_trans_2) # merge resulting rasters to create 0-360 inversed wind map
wind_dir_trans[wind_dir_trans < 1] = 1 # clean up the values between teh 360-1 degree boundary

#Extract a list of cells for the analysis from the watershed shapefile
sample.list = unlist(cellFromPolygon(elev,sample.poly))
```
##### *Ex. of GIS data used in the watershed source model*
```{r Plot Data, echo=FALSE}
plot(elev, main = "Elevation (m) with watershed polygon")
plot(sample.poly, add = TRUE)
plot(wind_dir_trans, main = "Wind direction (degrees) with watershed polygon")
plot(sample.poly, add = TRUE)
plot(wind_vel, main = "Wind velocity (mph) with watershed polygon")
plot(sample.poly, add = TRUE)
```


### Dispersion and Source Model


##### **User input variables needed for the script**

Variable Name  | Description                                 | Value    | Reference                            
-------------- | ------------------------------------------- | -------- | ---------------------------------
h              | Plume height (m)                            | 10       | Clark 1988; Peters & Higuera 2007
d              | Particle size (μm)                          | 250      | Clark 1988; Peters & Higuera 2007


##### **Constants needed for the script**

Constant Name  | Description                                 | Value    | Reference                            
-------------- | ------------------------------------------- | -------- | ---------------------------------
Cy             | Diffusion Constant                          | 0.21     | Peters & Higuera 2007; Sutton 1947
Cz             | Diffusion Constant                          | 0.12     | Peters & Higuera 2007; Sutton 1947
n              | Turbulence Measure                          | 0.25     | Sutton 1947
m              | n/(4−2n)                                    | --       | Peters & Higuera 2007
Qo             | Source strength                             | 100000   | Peters & Higuera 2007
pp             | Avg Particle density for charcoal (g/cm^-3) | 0.5      | Clark 1988; Sutton 1947
pf             | Fluid density for air (g/cm^-3 )            | 0.00127  | Clark 1988; Sutton 1947
v              | Kinematic viscosity for air (cm^2/sec^-1)   | 0.142    | Clark 1988; Sutton 1947
g              | Accel. due to gravity (cm/sec^-2)           | 981      | Clark 1988; Sutton 1947


##### *Complete code for dispersion and source model function*

```{r Watershed Source Model Function}
watershed.source.model = function(sample.list){
  
#######Create sample location and elevation
sample_num <-sample.list
sample = data.frame(xyFromCell(elev, sample_num))
sample <-c(sample[,1], sample[,2]) 
sample_elev = extract(elev, sample_num)


#### Create Distance raster
dist_rast <- distanceFromPoints(elev, sample)
dist_rast = dist_rast

#### Create Wind Direction Raster
elev_NA = elev
elev_NA[] = NA
elev_NA[sample_num]= 1 
# sample_num is the cell number for the sample location that was identified in the "import and create fire" section (for wtaershed analysis, this includes all teh cells located within teh drainage basin)
azimuth_rast=direction(elev_NA, degrees=TRUE, from = TRUE)

### Wind direction and difference raster
wind_diff <- (azimuth_rast)-(wind_dir_trans)
wind_diff_rads <- (wind_diff)*(pi/180)
wind_y_adj <- overlay(dist_rast, wind_diff_rads, fun=function(x,y){return(sqrt(((x^2)+(x^2))-((2)*(x*x)*cos(y))))})
wind_y_adj <- wind_y_adj 

### Plume elevation adjustment  raster
plume <- function(x) { x - sample_elev }
plume_adj <- calc(elev, plume)
#extract resolution for cell calculations
res = (xres(elev))

## User defined variables
h = 10
#u =  1 # user defined wind speed (m per sec ^-1 )
d =  250  #user defined diameter of particle in microns
#y = 0 # user defined to create cross section for y axis

## Constants
Cy = 0.21
Cz = 0.12
n = 0.25
m = n/(4-(2 * n))
Qo = 100000
e = exp(1)
pp = 0.5 # particle density for avg charcoal particle (g cm^-3) 
pf = 0.00127 # Fluid density for air (g cm^-3 )
v = 0.142 # kinematic viscosity for air (cm^2 sec^-1) 
g = 981 # Accel. due to gravity (cm sec^-2).

######### Settling Velocity equation (Stokes' Law)
settle.vel = function(r){
  dcent = 0.0001 * d
  vg = ((pp-pf)*g*(dcent^2))/(18*v)
  return(vg)
}

vg = settle.vel(r)

######### Charcoal model equation for x
disperse.x.y.elev.u = function(x,y,elev,u){
  
  ##Gamma function linear equivalent
  f = function(t) {exp(-t)*(t^((-m)-1))}
  xi = ((h + elev)^2)/(x^(2-n)*(Cz^2))
  z = integrate(f, lower = xi, upper = 10000)
  
  ## Qx function
  Qx = Qo*(exp(((4*vg)/(n*u*Cz*(sqrt(pi)))) * (((-x^(n/2))*(exp(-xi))) + (((h + elev)/Cz)^(2*m)) * (-m*z$value))))
  
  ## Dipsersal function
  dispersal = (((2*vg*Qx)/(u*pi*Cy*Cz*(x^(2-n))))*(exp((-y^2)/((Cy^2)*(x^(2-n)))))*(exp((-(h + elev)^2)/((Cz^2)*(x^(2-n))))))
  return(dispersal)
}

### Source area 
source_area <- overlay(dist_rast, wind_y_adj, plume_adj, wind_vel, fun=disperse.x.y.elev.u)
source_area[is.na(source_area)] = 0


res_source_area = overlay(source_area, fun =function(x){x*res*res})
sum_source_area = cellStats(res_source_area, sum)
percent_source_area <- overlay(res_source_area, fun =function(x){x/sum_source_area})

#Return the raster
returnlist = list(percent_source_area)
}
```

### Run Model and Process the Results
##### *Command to run model in parallel*
```{r Run Model}
# Run watershed.source.model
num.cores = detectCores()
result.nested = mclapply(sample.list, function(i){watershed.source.model(i)}, mc.cores = num.cores)
```

##### *Process results for final map*
```{r Process Results}
#Unlist rasters from results and calculate percent likelihood for the source area
watershed.source.model.result = unlist(result.nested)
watershed.source.model.sum = calc(brick(watershed.source.model.result[1:length(watershed.source.model.result)]),sum)
count.total = count(sample.list)
watershed.percent.source.area = (watershed.source.model.sum/(sum(count.total[,2])))

###Select cut-offs for displaying map (ie 30%, 50%, or standard deviations)
values_list = as.data.frame(watershed.percent.source.area)
values_list_descending = sort(values_list$layer, decreasing = TRUE)
cumsum_list = cumsum(values_list_descending)

#This example will use standard deviations for each cut-off
###Quantile 1 standard deviation 
per_1 <- which(cumsum_list < .68) # this is an adjustable percantage of the variation that the user would like to map
last_1 = per_1[length(per_1)]

###Quantile 2 standard deviations
per_2 <- which(cumsum_list < .95) # this is an adjustable percantage of the variation that the user would like to map
last_2 = per_2[length(per_2)]

###Quantile 3 standard deviations
per_3 <- which(cumsum_list < .997) # this is an adjustable percantage of the variation that the user would like to map
last_3 = per_3[length(per_3)]

###Group data by cut-off limits
cut_off_1 = values_list_descending[last_1]
cut_off_2 = values_list_descending[last_2]
cut_off_3 = values_list_descending[last_3]

##Code values in raster to reflect tehir cutt-off groups
watershed.percent.source.area.final = watershed.percent.source.area
watershed.percent.source.area.final[watershed.percent.source.area.final >= cut_off_1 & watershed.percent.source.area.final > 0] = 3
watershed.percent.source.area.final[watershed.percent.source.area.final >= cut_off_2 & watershed.percent.source.area.final < cut_off_1] = 2
watershed.percent.source.area.final[watershed.percent.source.area.final >= cut_off_3 & watershed.percent.source.area.final < cut_off_2 ] = 1
```

### Plot and save final result

```{r}
plot(watershed.percent.source.area.final)
#writeRaster(watershed.percent.source.area.final, filename="Path to save", format="GTiff", overwrite=TRUE)
```
### A prettier plot
```{r, ,message=FALSE}
#Convert raster results to shapefiles, re-project the shapefiles into lat/long WGS84, and fortify results for plotting

#Source areas as polygons
source.areas <- rasterToPolygons(watershed.percent.source.area.final, fun=function(x){x>=1}, dissolve = F)
source.areas <- spTransform(source.areas, CRS("+proj=longlat +datum=WGS84"))
source.areas.fort = fortify(source.areas, region = "layer")

#Transform Watershed shapefile
sample.poly.lat.long =  spTransform(sample.poly, CRS("+proj=longlat +datum=WGS84"))

#set up basemaps from Google Maps
Navarres_region <- get_map(location=c(lon = -0.725, lat = 39.09), zoom = 13, source="google", maptype="terrain",  crop = T) 

#Create plot using ggmap
ggmap(Navarres_region) +
  geom_polygon(aes(x=long, y=lat, group=group),data = sample.poly.lat.long, color ="black", size = .1, alpha = 0) +
  geom_polygon(aes(x=long, y=lat, fill = id, group=group), data = source.areas.fort, alpha = .6) +
  xlab("Longitude") +
  ylab("Latitude") +
  ggtitle("Charcoal Source Area Model\nWatershed v.1") +
  scale_fill_discrete(name="Source Area\nLikelihood",
                      breaks=c("1", "2", "3"),
                      labels=c("99%", "95%", "68%"))
```












